
---
title: "Homework Batch 2"
author: "Alyssa Gurkas, Deirdre Flynn, Marc Fridson"
output: 
  word_document:
    reference_docx: cerulean-style.docx
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,         # Show R code in the document
  results = 'markup',  # Show R output
  warning = FALSE,     # Suppress warning messages
  error = FALSE,       # Suppress error messages
  message = FALSE,     # Suppress messages
  fig.align = "center" # Center-align all figures
)

# Set theme
ggplot2::theme_set(ggplot2::theme_minimal() + 
  ggplot2::theme(plot.title = ggplot2::element_text(size = 14, face = "bold"),
        axis.title = ggplot2::element_text(size = 12)))

# insert libraries used here
```

# KJ 6.3
*Instructions*: A chemical manufacturing process for a pharmaceutical product was
discussed in Sect. 1.4. In this problem, the objective is to understand the 
relationship between biological measurements of the raw materials (predictors),
measurements of the manufacturing process (predictors), and the response of
product yield. Biological predictors cannot be changed but can be used to
assess the quality of the raw material before processing. On the other hand,
manufacturing process predictors can be changed in the manufacturing process.
Improving product yield by 1% will boost revenue by approximately
one hundred thousand dollars per batch: 

## KJ 6.3(a)
*Instructions*: Start R and use these commands to load the data:
> library(AppliedPredictiveModeling)
> data(chemicalManufacturing)

The matrix processPredictors contains the 57 predictors (12 describing
the input biological material and 45 describing the process predictors)
for the 176 manufacturing runs. yield contains the percent yield for each
run. 

### Approach - KJ 6.3(a)
[insert approach explanation here].

### Analysis - KJ 6.3(a)

```{r kj-6.3.a.-analysis}
# insert analysis here
```

## KJ 6.3(b)
*Instructions*: A small percentage of cells in the predictor set contain missing 
values. Use an imputation function to fill in these missing values 
(e.g., see Sect. 3.8).

### Approach - KJ 6.3(b)
[insert approach explanation here].

### Analysis - KJ 6.3(b)

```{r kj-6.3.b.-analysis}
# insert analysis here
```

## KJ 6.3(c)
*Instructions*: Split the data into a training and a test set, pre-process the 
data, and tune a model of your choice from this chapter. What is the optimal value
of the performance metric?

### Approach - KJ 6.3(c)
[insert approach explanation here].

### Analysis - KJ 6.3(c)

```{r kj-6.3.c.-analysis}
# insert analysis here
```

## KJ 6.3(d)
*Instructions*: Predict the response for the test set. What is the value of the 
performance metric and how does this compare with the resampled performance metric
on the training set?

### Approach - KJ 6.3(d)
[insert approach explanation here].

### Analysis - KJ 6.3(d)

```{r kj-6.3.d.-analysis}
# insert analysis here
```

## KJ 6.3(e)
*Instructions*: Which predictors are most important in the model you have trained? 
Do either the biological or process predictors dominate the list?

### Approach - KJ 6.3(e)
[insert approach explanation here].

### Analysis - KJ 6.3(e)

```{r kj-6.3.e.-analysis}
# insert analysis here
```

## KJ 6.3(f)
*Instructions*: Explore the relationships between each of the top predictors and the response. How could this information be helpful in improving yield in future runs 
of the manufacturing process?

### Approach - KJ 6.3(f)
[insert approach explanation here].

### Analysis - KJ 6.3(f)

```{r kj-6.3.f.-analysis}
# insert analysis here
```

# KJ 7.2 
*Instructions*: Which models appear to give the best performance? Does MARS select the informative predictors (those named X1–X5)?

### Approach - KJ 7.2
[insert approach explanation here].

### Analysis - KJ 7.2

```{r kj-7.2-analysis}
# insert analysis here
```

# KJ 7.5 
*Instructions*: Exercise 6.3 describes data for a chemical manufacturing process. Use the same data imputation, data splitting, and pre-processing steps as before and train several nonlinear regression models.

## KJ 7.5(a) 
*Instructions*: Which nonlinear regression model gives the optimal resampling and test set performance?

### Approach 
[insert approach explanation here].

### Analysis - KJ 7.5(a)

```{r kj-7.2.a.-analysis}
# insert analysis here
```

## KJ 7.5(b) 
*Instructions*: Which predictors are most important in the optimal nonlinear regression model? Do either the biological or process variables dominate the list? How do the top ten important predictors compare to the top ten predictors from the optimal linear model?

### Approach - KJ 7.5(b)
[insert approach explanation here].

### Analysis - KJ 7.5(b)

```{r kj-7.2.b.-analysis}
# insert analysis here
```

## KJ 7.5(c) 
*Instructions*: Explore the relationships between the top predictors and the response for the predictors that are unique to the optimal nonlinear regression model. Do these plots reveal intuition about the biological or process predictors and their relationship with yield?

### Approach - KJ 7.5(c)
[insert approach explaination here].

### Analysis - KJ 7.5(c)

```{r kj-7.2.c.-analysis}
# insert analysis here
```

# KJ 8.1 Recreate the simulated data from Exercise 7.2.

## KJ 8.1(a) 
*Instructions*: Fit a random forest model to all of the predictors, then estimate the variable importance scores. Did the random forest model significantly use the uninformative predictors (V6 – V10)?

### Approach - KJ 8.1(a)
[insert approach explaination here].

### Analysis - KJ 8.1(a)

```{r kj-8.1.a.-analysis}
# insert analysis here
```

## KJ 8.1(b) 
*Instructions*: Now add an additional predictor that is highly correlated with one of the informative predictors. Fit another random forest model to these data. Did the importance score for V1 change? What happens when you add another predictor that is also highly correlated with V1?

### Approach - KJ 8.1(b)
[insert approach explaination here].

### Analysis - KJ 8.1(b)

```{r kj-8.1.b.-analysis}
# insert analysis here
```

## KJ 8.1(c) 
*Instructions*: Use the cforest function in the party package to fit a random forest model using conditional inference trees. The party package function varimp can calculate predictor importance. The conditional argument of that function toggles between the traditional importance measure and the modified version described in Strobl et al. (2007). Do these importances show the same pattern as the traditional random forest model?

### Approach - KJ 8.1(c)
[insert approach explaination here].

### Analysis - KJ 8.1(c)

```{r kj-8.1.c.-analysis}
# insert analysis here
```

## KJ 8.1(d) 
*Instructions*: Repeat this process with different tree models, such as boosted 
trees and Cubist. Does the same pattern occur?

### Approach - KJ 8.1(d)
[insert approach explaination here].

### Analysis - KJ 8.1(d)

```{r kj-8.1.d.-analysis}
# insert analysis here
```

# KJ 8.2 
*Instructions*: Use a simulation to show tree bias with different granularities.

### Approach - KJ 8.2
[insert approach explaination here].

### Analysis - KJ 8.2

```{r kj-8.2.-analysis}
# insert analysis here
```

# KJ 8.3 
*Instructions*: In stochastic gradient boosting the bagging fraction and learning rate will govern the construction of the trees as they are guided by the gradient. Although the optimal values of these parameters should be obtained through the tuning process, it is helpful to understand how the magnitudes of these parameters affect magnitudes of variable importance. Figure 8.24 provides the variable importance plots for boosting using two extreme values for the bagging fraction (0.1 and 0.9) and the learning rate (0.1 and 0.9) for the solubility data. The left-hand plot has both parameters set to 0.1, and the right-hand plot has both set to 0.9:

## KJ 8.3(a)
*Instructions*: Why does the model on the right focus its importance on just the 
first few of predictors, whereas the model on the left spreads importance across
more predictors?

### Approach - KJ 8.3(a)
[insert approach explaination here].

### Analysis - KJ 8.3(a)

```{r kj-8.3.a.-analysis}
# insert analysis here
```

## KJ 8.3(b)
*Instructions*: Which model do you think would be more predictive of other 
samples?

### Approach - KJ 8.3(b)
[insert approach explaination here].

### Analysis - KJ 8.3(b)

```{r kj-8.3.b.-analysis}
# insert analysis here
```

## KJ 8.3(c)
*Instructions*: How would increasing interaction depth affect the slope of 
predictor importance for either model in Fig. 8.24?

### Approach - KJ 8.3(c)
[insert approach explaination here].

### Analysis - KJ 8.3(c)

```{r kj-8.3.c.-analysis}
# insert analysis here
```

# KJ 8.4
*Instructions*: Use a single predictor in the solubility data, such as the 
molecular weight or the number of carbon atoms and fit several models. Plot the 
predictor data versus the solubility results for the test set. Overlay the model 
predictions for the test set. How do the model differ? Does changing the tuning 
parameter(s) significantly affect the model fit?

## KJ 8.4(a)

### Approach - KJ 8.4(a)
[insert approach explaination here].

### Analysis - KJ 8.4(a)

```{r kj-8.4.a.-analysis}
# insert analysis here
```

## KJ 8.4(b)

### Approach - KJ 8.4(b)
[insert approach explaination here].

### Analysis - KJ 8.4(b)

```{r kj-8.4.b.-analysis}
# insert analysis here
```

## KJ 8.4(c)

### Approach - KJ 8.4(c)
[insert approach explaination here].

### Analysis - KJ 8.4(c)

```{r kj-8.4.c.-analysis}
# insert analysis here
```

# KJ 8.7
*Instructions*: Refer to Exercises 6.3 and 7.5 which describe a chemical 
manufacturing process. Use the same data imputation, data splitting, and 
pre-processing steps as before and train several tree-based models:

## KJ 8.7(a)

### Approach - KJ 8.7(a)
[insert approach explaination here].

### Analysis - KJ 8.7(a)

```{r kj-8.7.a.-analysis}
# insert analysis here
```

## KJ 8.7(b)
*Instructions*: Which predictors are most important in the optimal tree-based 
regression model? Do either the biological or process variables dominate the list?
How do the top 10 important predictors compare to the top 10 predictors from the 
optimal linear and nonlinear models?

### Approach - KJ 8.7(b)
[insert approach explaination here].

### Analysis - KJ 8.7(b)

```{r kj-8.7.b.-analysis}
# insert analysis here
```

## KJ 8.7(c)
*Instructions*: Plot the optimal single tree with the distribution of yield in 
the terminal nodes. Does this view of the data provide additional knowledge about
the biological or process predictors and their relationship with yield?

### Approach - KJ 8.7(c)
[insert approach explaination here].

### Analysis - KJ 8.7(c)

```{r kj-8.7.c.-analysis}
# insert analysis here
```
