---
title: "Week 1  Homework"
author: "Alyssa Gurkas"
date: "2025-05-31"
output: html_document
---
## Homework Exercises from Forecasting: Principles and Practice
### Exercise 2.1 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-libraries}
library(fpp2)
library(tidyverse)
```

### 1. Use the help function to explore what the series gold, woolyrnq and gas represent.

```{r help-gold}
help("gold")
```

The gold dataset represents the daily gold prices in US dollars from January 1, 1985
to March 31, 1989.

```{r help-woolyrnq}
help("woolyrnq")
```

The woolyrnq dataset represents the quarterly Australian production of woollen 
yarn (in tonnes) from March 1965 - September 1994. 

```{r help-gas}
help("gas")
```

The gas dataset represents the Australian monthly gas production from 1956-1995.

### 1.a. Use autoplot() to plot each of these in separate plots.

#### `gold` Plot

```{r gold-plot fig.cap = "Figure 1. Daily Gold Prices in US dollars: January 1, 1985 to March 31, 1989"}
autoplot(gold)+
  xlab("Time") + ylab("U.S. Dollars")
```


#### `woolyrnq` Plot
```{r woolyrnq-plot, fig.cap="Figure 2. Quarterly Australian Production of Woollen Yarn (in tonnes) from March 1965 - September 1994"}
autoplot(woolyrnq)+
  xlab("Time") + ylab("Woollen Yarn Production (Tonnes)")
```

#### `gas` Plot
```{r gas-plot, fig.cap= "Figure 3. Australian Monthly Gas Production from 1956-1995"}
autoplot(woolyrnq)+
  xlab("Time") + ylab("Gas Production")
```

### 1.b. What is the frequency of each series? Hint: apply the frequency() function.
Gold Frequency: `r stats::frequency(gold)`
Woollen Yarn Frequency: `r stats::frequency(woolyrnq)`
Gas Frequency: `r stats::frequency(gas)`

```{r freq}
frequency(gold)
frequency(woolyrnq)
frequency(gas)
```


### 1.c. Use which.max() to spot the outlier in the gold series. Which observation was it?
```{r outlier-gold}
which.max(gold)
```

The outlier in the gold series is `r which.max(gold)` U.S. Dollars.

## 2. Download the file tute1.csv from the book website, open it in Excel (or some other spreadsheet application), and review its contents. You should find four columns of information. Columns B through D each contain a quarterly series, labelled Sales, AdBudget and GDP. Sales contains the quarterly sales for a small company over the period 1981-2005. AdBudget is the advertising budget and GDP is the gross domestic product. All series have been adjusted for inflation.

```{r load-tute1}


```

